---
title: "R Code for Chapter 5 of Introduction to Data Mining: Association Rule Mining (Additional Interactive Visualization)"
author: "Michael Hahsler"
output:
  html_document:
    toc: true
---

This code covers chapter 5 of "Introduction to Data Mining" by Pang-Ning Tan, Michael Steinbach and Vipin Kumar. This work is licensed under the Creative Commons Attribution 4.0 International License. For questions please contact Michael Hahsler.

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(arules)
library(arulesViz)
```

Load the data set

```{r}
data(Groceries)
summary(Groceries)
inspect(head(Groceries))
```

Mine Association Rules

```{r}
rules <- apriori(Groceries, parameter=list(support=0.001, confidence=.8))
```

## Interactive inspect with sorting, filtering and paging
```{r}
inspectDT(rules)
```

## Plot rules as a scatter plot using an interactive html widget
```{r}
plot(rules, engine = "html")
```

**Note**: plotly currently does not do well with too many points, so plot selects the top 1000 rules only (with a warning).

## Plot rules as a graph using an interactive html widget
```{r}
plot(rules, method = "graph", engine = "html")
```

**Note**: currently does not do well with too many points, so plot selects the top 100 rules only (with a warning).

## Interactive Rule Explorer

To use the interactive Rule Explorer, paste this into your R session:
```{r}
library("arulesViz")
data(Groceries)
# First, mine the rules
rules <- apriori(Groceries, parameter = list(support = 0.001, confidence = 0.8))

# Then, use ruleExplorer on the mined rules
ruleExplorer(rules)
```
