---
title: "R Code for Chapter 5 of Introduction to Data Mining: Association Analysis"
author: "Michael Hahsler"
output:
  html_document:
    toc: true
---

This code covers chapter 5 of "Introduction to Data Mining" by Pang-Ning Tan, Michael Steinbach and Vipin Kumar.

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Used packages

Install the following packages:

* arules
* arulesViz

```{r}
library(arules)
library(arulesViz)
```

# Used data
```{r}
data(Zoo, package="mlbench")
head(Zoo)
```

# Transactions
## Create transactions
```{r}
try(trans <- as(Zoo, "transactions"))
```

What is column 13?

```{r}
colnames(Zoo)[13]
legs <- Zoo[["legs"]]
summary(legs)
hist(legs)
table(legs)
```

Make legs into has/does not have legs:
```{r}
has_legs <- legs>0
has_legs
table(has_legs)
Zoo[["legs"]] <- has_legs

# Convert data into a set of transactions
trans <- as(Zoo, "transactions")
trans
```

## Inspect Transactions
```{r}
summary(trans)

# Look at created items
colnames(trans)

# Compare with the original features from Zoo
colnames(Zoo)

# Look at a few transactions as a matrix
as(trans, "matrix")[1:3,]

# Look at the transactions as sets of items
inspect(trans[1:3])

# Plot the binary matrix
image(trans)

# Look at the relative frequency of items
itemFrequencyPlot(trans,topN=20)
plot(sort(itemFrequency(trans, type="absolute"), decreasing=TRUE),
  xlab = "Items", ylab="Support Count", type="l")
```

## Alternative encoding
```{r}
Zoo2 <- Zoo
for(i in 1:ncol(Zoo2)) Zoo2[[i]] <- as.factor(Zoo2[[i]])
sapply(Zoo2, class)
summary(Zoo2)

trans2 <- as(Zoo2, "transactions")
trans2

itemFrequencyPlot(trans2, topN=20)

# Select transactions that contain a certain item
trans_insects <- trans2[trans %in% "type=insect"]
trans_insects
inspect(trans_insects)
```

## Vertical layout (Transaction ID Lists)
```{r}
vertical <- as(trans, "tidLists")
as(vertical, "matrix")[1:10,1:5]
```

# Frequent Itemsets
## Mine Frequent Itemsets
```{r}
2^ncol(trans)

is <- apriori(trans, parameter=list(target="frequent"))
is

5/nrow(trans)

is <- apriori(trans, parameter=list(target="frequent", support=0.05))
is

# Sort by support
is <- sort(is, by="support")
inspect(head(is, n=10))

# Look at frequent itemsets with many items
barplot(table(size(is)), xlab="itemset size", ylab="count")
inspect(is[size(is)>8])
```

## Concise Representation of Itemsets

```{r}
# Find maximal frequent itemsets
is_max <- is[is.maximal(is)]
is_max
inspect(head(sort(is_max, by="support")))

# Find closed frequent itemsets
is_closed <- is[is.closed(is)]
is_closed
inspect(head(sort(is_closed, by="support")))

barplot(c(
  frequent=length(is),
  closed=length(is_closed),
  maximal=length(is_max)
  ), ylab="count", xlab="itemsets")
```

# Association Rules
## Mine Association Rules
```{r}
rules <- apriori(trans, parameter=list(support=0.05, confidence=.9))
length(rules)

inspect(head(rules))
quality(head(rules))

# Look at rules with highest lift
rules <- sort(rules, by="lift")
inspect(head(rules, n=10))

# Create rules using the alternative encoding
r <- apriori(trans2)
r
print(object.size(r), unit="Mb")

inspect(r[1:10])
inspect(head(r, by="lift", n = 10))
```

## Additional Interest Measures
```{r}
interestMeasure(rules[1:10], measure=c("phi", "gini"),
  trans=trans)

# Add measures to the rules
quality(rules) <- cbind(quality(rules),
  interestMeasure(rules, measure=c("phi", "gini"),
    trans=trans))

# Find rules which score high for Phi correlation
inspect(head(rules, by="phi"))
```

## Mine using Templates
```{r}
type <- grep("type=", itemLabels(trans), value = TRUE)
type

rules_type <- apriori(trans,
  appearance= list(rhs=type, default="lhs"))

inspect(head(sort(rules_type, by="lift")))
```

## Association rule visualization
```{r}
# Default scatterplot
plot(rules)

# Add some jitter
plot(rules, control=list(jitter=.5))

plot(rules, shading="order", control=list(jitter=.5))

# Grouped plot
plot(rules, method="grouped")

# As a graph
plot(rules, method="graph")
plot(head(rules, by="phi", n = 100), method="graph")
```
